---
title: "Practical Machine Learning Course Project"
output: html_document
---

##Executive Summary
Human Activity Recognition (HAR) is a key research area that is gaining increasing attention, especially for the development of context-aware systems. There are many potential applications for HAR, like: elderly monitoring, life log systems for monitoring energy expenditure and for supporting weight-loss programs, and digital assistants for weight lifting exercises. Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.

Six participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 
(Class A), exactly according to the specification. 
(Class B), throwing the elbows to the front. 
(Class C), lifting the dumbbell only halfway. 
(Class D), lowering the dumbbell only halfway. 
(Class E) and throwing the hips to the front.

This report will describe how the data captured are used to identify the parameters involved in predicting the movement involved based on the above classification, and then to predict the movement for 20 test cases.

The training data were divided into two groups, a training data and a validation data (to be used to validate the data), to derived the prediction model by using the training data, to validate the model where an expected out-of-sample error rate of less than 0.5%, or 99.5% accuracy, would be acceptable before it is used to perform the prediction on the 20 test cases - that must have 100% accuracy (to obtain 20 points awarded).

The training model developed using Random Forest was able to achieve over 99.99% accuracy, or less than 0.03% out-of-sample error, and was able to predict the 20 test cases with 100% accuracy.


##Reproduceablity
In order to reproduce the same results, you need a certain set of packages, as well as setting a pseudo random seed equal to the one I used. *Note:To install, for instance, the caret package in R, run this command: install.packages(“caret”)

The following Libraries were used for this project, which you should install - if not done yet - and load on your working environment.

```{r}
library(caret)
```


```{r}
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
```

```{r}
library(randomForest)
```

Finally, load the same seed with the following line of code:

```{r}
set.seed(12345)
```

##Getting the data
The training data set can be found on the following URL:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```

The testing data set can be found on the following URL:

```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Load data to memory solely

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

##Partioning the training set into two
Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:
```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```

##Cleaning the data
The following transformations were used to clean the data:

Transformation 1: Cleaning NearZeroVariance Variables Run this code to view possible NZV Variables:

```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```
Run this code to create another subset without NZV variables:

```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
#To check the new N?? of observations
dim(myTraining)
```

Transformation 2: Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms:

```{r}
myTraining <- myTraining[c(-1)]
```

Transformation 3: Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA’s I’m going to leave them out:

```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}
#To check the new N?? of observations
dim(trainingV3)
```

```{r}
#Seting back to our set:
myTraining <- trainingV3
rm(trainingV3)
```

Now let us do the exact same 3 transformations but for our myTesting and testing data sets.

```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]

#To check the new N?? of observations
dim(myTesting)
```


```{r}
#To check the new N?? of observations
dim(testing)
```
In order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
#And to make sure Coertion really worked, simple smart ass technique:
testing <- rbind(myTraining[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```

##Using ML algorithms for prediction: Decision Tree

```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```

Note: to view the decision tree with fancy run this command:

```{r}
fancyRpartPlot(modFitA1)
```

Predicting:

```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

(Moment of truth) Using confusion Matrix to test results:

```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

##Using ML algorithms for prediction: Random Forests

```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```

Predicting in-sample error::

```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```

(Moment of truth) Using confusion Matrix to test results:

```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```

Random Forests yielded better Results, as expected!

##Generating Files to submit as answers for the Assignment:
Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which yielded a much better prediction in in-sample:

```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```

Function to generate files with predictions to submit for assignment

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```

Apply the prediction model to the testing data. The predicted classification are (and were 100% accurate):

```{r}
predictionsB2
```


##Conclusion
The model predicted the 20 test cases with 100% accuracy. All 20 points were awarded after submitting the 20 test files.

## End
